### 2022-11-15 NSF2022/Description/broaderimpacts.tex
changes throughout

\subsection{Educational Impacts}
\label{subsec:Educational_plans}
To foster training in mathematical sciences,
the proposal supports two undergraduate researchers
(URs) per year at Fordham University---for eight summer weeks.
The URs will collaborate directly with the PIs,
receive tutorials e.g., numerical methods,
and be guided on mathematical writing and presentation.
Fordham Mathematics recently completed a collaboration studio 
purpose-built for UR support.
Work products include a summary manuscript,
a presentation in the Fordham Undergraduate Research Symposium
and departmental seminars, and participation in a national conference.
We have outlined three projects for the present proposal.
\begin{enumerate}[noitemsep,topsep=0pt]
\item Translate project code from MATLAB into a compiled language 
\item Use machine learning and visualizations to post-process data of granule dynamics
\item Mathematical modeling of group affiliation in social science  
\end{enumerate}
PI Ryham has worked side-by-side with dozens of
undergraduates and high-schoolers in
supported summer research, Girls in Science Technology and Mathematics,
and the Clare Boothe Luce Scholarship program
and the activity benefited their careers in
machine learning, water resources engineering, threat detection,
and the NYC Department of Education.
Presently PI Ryham is mentoring a Clare Boothe Luce Scholar on
polling accessibility in New York City.
The PI is finishing works on energy densities for lipid  bilayers
with a recently graduated UR, now working at Raytheon Technologies.
His publications include works
with seven undegraduate coauthors
~\cite{Figueroa2012CuttingCI, RYHAM20112929, RyWaCo13, RyKlYaCo16}.

To increase the participation of women,
persons with disabilities and underrepresented minorities in STEM,
the PIs will specifically target students
whose socio-economic background prevents them from participating in
out-of-state research experiences.
The inclusion objectives will be aided by the fact that  
Fordham's class of 2026 weighs in with 63 \% women
and more than 46 \% domestic students of color. 
PI Ryham will solicit applicants from
Fordham University's Collegiate Science and Technology Entry Program.
Students from the Bronx High School of Science (PI Ryham) and Newark
Science Park High School (PI Young) will also be encouraged to join the
research team.

PI Young has been actively involved with promotion of
underrepresented students at NJIT.  A former PhD student, Herve Nganguia, is African American and 
now an Assistant Professor at Towson University. 
%PI Young has taught a
%broad spectrum of courses in fluid mechanics and applied math modeling.
PI Quaife will continue to work with undergraduates and high school
students through the Undergraduate Research Opportunity Program and
Young Scholar's Program.
%, as he has done in the past.

%To maximize vertical integration, the PIs and personnel will travel to
%New York for one or two weeks during the summers. 
Finally, we will
create several modules to include in our courses. These modules will
introduce topics from numerical linear algebra and optimization and
highlight concrete applications of mathematics based on the ideas of the
proposal.

>>>>

\subsection{Educational Impacts}
\label{subsec:Educational_plans}
To foster training in mathematical sciences, the proposal supports two
undergraduate researchers (URs) per year at Fordham University---for
eight summer weeks. The URs will collaborate directly with the PIs,
receive tutorials (e.g., numerical methods), and be guided on
mathematical writing and presentation. Fordham Mathematics recently
completed a collaboration studio purpose-built for UR support. Work
products include a summary manuscript, a presentation in the Fordham
Undergraduate Research Symposium and departmental seminars, and
participation in a national conference. We have outlined three projects
for the present proposal:
\begin{enumerate}[noitemsep,topsep=0pt]
  \item Translate project code from MATLAB into a compiled language;
  \item Use machine learning and visualizations to post-process data of
    granule dynamics;
  \item Mathematical modeling of group affiliation in social science.
\end{enumerate}
PI Ryham has worked side-by-side with dozens of undergraduates and
high-schoolers in supported summer research, Girls in Science Technology
and Mathematics, and the Clare Boothe Luce Scholarship program and the
activity benefited their careers in machine learning, water resources
engineering, threat detection, and the NYC Department of Education.
Presently PI Ryham is mentoring a Clare Boothe Luce Scholar on polling
accessibility in New York City. The PI is finishing works on energy
densities for lipid bilayers with a recently graduated UR, now working
at Raytheon Technologies. His publications include works with seven
undergraduate coauthors~\cite{Figueroa2012CuttingCI, RYHAM20112929,
RyWaCo13, RyKlYaCo16}.

To increase the participation of women, persons with disabilities and
underrepresented minorities in STEM, the PIs will specifically target
students whose socio-economic background prevents them from
participating in out-of-state research experiences. The inclusion
objectives will be aided by the fact that Fordham's class of 2026 weighs
in with 63\% women and more than 46\% domestic students of color. PI
Ryham will solicit applicants from Fordham University's Collegiate
Science and Technology Entry Program. Students from the Bronx High
School of Science (PI Ryham) and Newark Science Park High School (PI
Young) will also be encouraged to join the research team.

PI Young has been actively involved with promotion of underrepresented
students at NJIT. A former PhD student, Herve Nganguia, is African
American and now an Assistant Professor at Towson University. PI Quaife
is currently advising three female PhD students, and one is African
American and the recipient of the McKnight Doctoral Fellowship which
addresses the underrepresentation of African American and Hispanic
faculty at colleges and universities in the state of Florida. PI Quaife
will also continue to work with undergraduates and high school students
through the Undergraduate Research Opportunity Program and Young
Scholar's Program.

Finally, we will create several modules to include in our courses. These
modules will introduce topics from numerical linear algebra and
optimization and highlight concrete applications of mathematics based on
the ideas of the proposal.

### 2022-11-15 NSF2022/Description/specificaim2.tex
changes throughout

While the inclusion of high-order quadrature and fast methods is
necessary to simulate dense suspensions, an additional required
computational tool is a collision-aware adaptive time stepping method.
The phase field results in the granules attracting one another and this
leads to nearly contacting boundaries that, even with high-order spatial
accuracy, can collide due to time stepping error. Our current
implementation uses a repulsive Lennard-Jones force, but this introduces
stiffness, thereby requiring small time steps, and this can be a
bottleneck when simulating
\begin{wrapfigure}[26]{l}{1.8in}
  \vspace{-0pt}
  \centering
  \input{figures/doublewell.tikz} 
  \vspace{-10pt}
  \caption{\label{fig:CA} \footnotesize (a) A double-well potential. (b)
  The solution of the one-dimensional phase
  transition~\eqref{eqn:phase}. (c) The convergence of exact and inexact
  Newton's methods.}
\end{wrapfigure}
dense suspensions. Therefore, we will develop a high-order
collision-avoiding adaptive time stepping method. This method will avoid
collisions by using the minimal separation distance to impose a
non-overlapping constraint. This results in a linear complementarity
problem that can be solve efficiently with a projected gradient descent
method~\cite{Yan2019}. High-order time adaptivity will be based on PI
Quaife's previous work on suspensions~\cite{qua-bir2016}, and this
method has proven to be a robust method to simulate dense
suspensions~\cite{qua-vee-you2019, kab-qua-bir2017, qua-gan-you2021}.
Following this previous work, the proposed work will use a spectral
deferred correction method that iteratively applies a low-order,
single-step method to achieve high-order accuracy. At each time step, we
will compute the total force and torque of the system which is
computationally cheap and physically zero, and therefore provides an
estimate of the time stepping error. PI Quaife's experience is that
error estimates based on physical constraints, such as force- and
torque-free conditions, appropriately determine adaptive time step sizes
so that the dynamics are resolved without the computational expense of
techniques such as embedded Runge-Kutta, step-doubling, and Richardson
extrapolation.

The phase field is governed by the PDE~\eqref{eqn:phase} in the solvent
region $\Omega(t)$. If $W(\phi) = \tfrac{1}{2}\phi^2$,
equation~\eqref{eqn:phase} is linear, and the PIs have used IEs to solve
address this problem~\cite{Fu2018_SIAM, FuQuRyYo22, fu-ryh-qua-you2022},
but for a double-well potential $W(\phi)$, the resulting PDE is
nonlinear, motivating the following problem:

@@

For a general double-well potential $W(\phi)$ with local minima at
$\phi_0$ and $\phi_1$, we propose to interpret the solution
of~\eqref{eqn:phase} as a root finding problem for
\begin{align}
  \label{eqn:phaseRoot}
  F[\phi] = -\epsilon^2 \Delta \phi + W'(\phi).
\end{align}
Equation~\eqref{eqn:phaseRoot} will be solved with Newton's method,
$\phi^{N+1} = \phi^{N} - \alpha (F'(\phi^N))^{-1} F(\phi^N)$, where
$F'(\phi^N) = -\epsilon^2 \Delta + W''(\phi^N)$ is the
Fr\'{e}chet derivative of $F$, and $\alpha$ is the step size that will
be determined using backtracking. This requires solving the variable
coefficient PDE $-\epsilon^2 \Delta v + W''(\phi^{N}) v =
F[\phi^N]$, which is challenging with an integral equation since a
fundamental solution is not readily available. However, by choosing a
double-well potential with $W''(\phi_0) = W''(\phi_1) = \beta$, we
propose to instead use an inexact Newton's method where $(F')^{-1}$ is
approximated with the solution of $-\epsilon^2 \Delta v +
\beta v = F[\phi^N]$. We test the efficacy of the exact and
inexact Newton's method applied to~\eqref{eqn:phaseRoot} in $(0,1)
\subset \RR$ by discretizing $F'$ with a second-order finite difference
stencil. The double-well potential is illustrated in
Figure~\ref{fig:CA}(a), and both the exact and inexact Newton's methods
converge to the solution in Figure~\ref{fig:CA}(b).
Figure~\ref{fig:CA}(c) shows the norm of the residual at each Newton
iterate when using the exact second derivative (red curve) and the
approximate second derivative (blue curve). While the inexact Newton's
method requires roughly twice as many iterations to reach a tolerance
of $10^{-4}$, each iteration is significantly cheaper to apply.

>>>>

While the inclusion of high-order quadrature and fast methods is
necessary to simulate dense suspensions, an additional required
computational tool is a collision-aware adaptive time stepping method.
The phase field results in the granules attracting one another and this
leads to nearly contacting boundaries that, even with high-order spatial
accuracy, can collide due to time stepping error. Our current
implementation uses a repulsive Lennard-Jones force, but this introduces
stiffness, thereby requiring small time steps, and this can be a
bottleneck when simulating
dense suspensions. Therefore, we will develop a high-order
collision-avoiding adaptive time stepping method. This method will avoid
collisions by using the minimal separation distance to impose a
non-overlapping constraint. This results in a linear complementarity
problem that can be solve efficiently with a projected gradient descent
method~\cite{Yan2019}. High-order time adaptivity will be based on PI
Quaife's previous work on suspensions~\cite{qua-bir2016}, and this
method has proven to be a robust method to simulate dense
suspensions~\cite{qua-vee-you2019, kab-qua-bir2017, qua-gan-you2021}.
Following this previous work, the proposed work will use a spectral
deferred correction method that iteratively applies a low-order,
single-step method to achieve high-order accuracy. At each time step, we
will compute the total force and torque of the system which is
computationally cheap and physically zero, and therefore provides an
estimate of the time stepping error. PI Quaife's experience is that
error estimates based on physical constraints, such as force- and
torque-free conditions, appropriately determine adaptive time step sizes
so that the dynamics are resolved without the computational expense of
techniques such as embedded Runge-Kutta, step-doubling, and Richardson
extrapolation.

The phase field is governed by the PDE~\eqref{eqn:phase} in the solvent
region $\Omega(t)$. If $W(\phi) = \tfrac{1}{2}\phi^2$,
equation~\eqref{eqn:phase} is linear, and the PIs have used IEs to solve
this problem~\cite{Fu2018_SIAM, FuQuRyYo22, fu-ryh-qua-you2022}, but for
a double-well potential $W(\phi)$, the resulting PDE is nonlinear,
motivating the following problem:

@@

For a general double-well potential $W(\phi)$ with local minima at
$\phi_0$ and $\phi_1$, we propose to interpret the solution
of~\eqref{eqn:phase} as a root finding problem for
\begin{align}
  \label{eqn:phaseRoot}
  F[\phi] = -\epsilon^2 \Delta \phi + W'(\phi).
\end{align}
Equation~\eqref{eqn:phaseRoot} will be solved with Newton's method,
$\phi^{N+1} = \phi^{N} - \alpha (F'(\phi^N))^{-1} F(\phi^N)$, where
$F'(\phi^N) = -\epsilon^2 \Delta + W''(\phi^N)$ is the Fr\'{e}chet
derivative of $F$, and $\alpha$ is the step size that will be determined
using backtracking. This requires solving the variable coefficient PDE
$-\epsilon^2 \Delta v + W''(\phi^{N}) v = F[\phi^N]$, which is
challenging with an integral equation since a fundamental solution is
not readily available. However, by choosing a double-well potential with
$W''(\phi_0) = W''(\phi_1) = \beta$, we propose to instead use an
inexact Newton's method where $(F')^{-1}$ is approximated with the
solution of $-\epsilon^2 \Delta v + \beta v = F[\phi^N]$. We test the
efficacy of the exact and inexact Newton's method applied
to~\eqref{eqn:phaseRoot} in $(0,1) \subset \RR$ by discretizing $F'$
with a second-order finite difference stencil. The double-well potential
is illustrated in Figure~\ref{fig:CA}(a), and both the exact and inexact
Newton's methods converge to the same solution. Figure~\ref{fig:CA}(b)
shows the norm of the residual at each Newton
\begin{wrapfigure}[18]{l}{1.8in}
  \vspace{-7pt}
  \centering
  \input{figures/doublewell.tikz} 
  \vspace{-10pt}
  \caption{\label{fig:CA} \footnotesize (a) A double-well potential. (b)
  The convergence of exact and inexact Newton's methods applied
  to~\eqref{eqn:phaseRoot}.}
\end{wrapfigure}
iterate when using the exact second derivative (red curve) and the
approximate second derivative (blue curve). While the inexact Newton's
method requires roughly twice as many iterations to reach a tolerance
of $10^{-4}$, each iteration is significantly cheaper to apply.
